Learning to Synthesize Rules for Network Intrusion Detectors

### 2020-02-22 paper/paper.tex
pass on 3.1 3.2
section 3.1

A
Suricata rule is divided in three parts---action, header, and rule
options~\cite{suri-rule-format}. The action part appears as the first
word in the rule description. An action denotes the task that needs to
be executed if the rule pattern is satisfied.
In this example, a
message will be sent to system administrators if the rule pattern is
satisfied. The header comes after the action in the rule
description. It restricts the information flow covered by the
rule. For this rule, the header is \CodeIn{tcp \$HOME\_NET any ->
  \$EXTERNAL\_NET any}. It instructs Suricata to inspect \CodeIn{tcp}
traffic flowing from any port in the home network to any other address
outside the home network. The variables \CodeIn{\$HOME\_NET} and
\CodeIn{\$EXTERNAL\_NET} are configurable. The rule options come after
the header. It is a semi-colon-separated sequence of key-value
pairs. Rule options serve to document the analyzed traffic (\eg{},
``msg'', ``classtype'', and ``sid'') and to characterize the attack
pattern (\eg, ``flags'', ``threshold'', and ``content''). \tname{}
produces options of the second kind, which are \emph{relevant} to
capturing the attack.

\vspace{-1ex}
\begin{table}[h!]
  \footnotesize
  \caption{\label{table:rules}Example options of Suricata.}
  \vspace{-2ex}
  \centering
  \begin{tabular}{clp{5.5cm}}
    \toprule
    \multicolumn{1}{c}{\#} & \multicolumn{1}{c}{Name} &  \multicolumn{1}{c}{Description}\\
    \midrule     
    1 & dsize & matches on the size of the packet payload\\
    2 & itype &  matches on a specific ICMP type\\
    3 & icode & matches on a specific ICMP code\\
    4 & icmp\_seq  & checks a ICMP sequence number\\
    5 & icmp\_id & matches on specific ICMP id-values\\
    6 & window & checks the size of the TCP window\\
    7 & flags & matches on TCP flags\\
    8 & fragbits & checks if the fragmentation or reserved bits are set in the IP header\\
    9 & threshold & controls the rule’s alert frequency\\
    10 & content & checks if argument is present on the payload\\
    \bottomrule
  \end{tabular}
\end{table}
%\vspace{-1ex}

@@

The main challenge in synthesizing rules is to discover the set of
options in a rule that blocks the malicious traffic, but allows the
benign traffic to pass. \tname{} proceeds as follows to synthesize
rules for this attack.

First, \tname{} analyzes the malicious traffic and extracts options
expressed in that traffic. Recall from
Section~\ref{sec:rules-and-packets} that rule options can be inferred
from packet contents. In this case, however, that information alone is
insufficient as the malicious traffic is characterized as a sequence
of 5K\Mar{really 5K?}\Mar{...and where did they come from?}
messages. \tname{} finds the following set of options at this stage by
analyzing the flow and contents of the messages:

@@

Finally, \tname{} ranks the inferred rules according to heuristics
learned from a database of existing rules. For example, one heuristic
function \tname{} uses for ranking leverages the amount of options a
rule of a given protocol typically has---the score of a rule on this
function is proportional to the popularity of the number of options it
declares. The list below shows the rule options for the top-5 rules
\tname{} produces for this attack.\Mar{algum problema em fazer isto?
  exemplos concretos ajudam muito!}

To sum up, \tname{} uses the messages that manifest an attack in the
first step to create an overspecific seed rule, benign traffic in the
second step to create a set of plausible rules from the seed rule, and
existing rules from popular sources in the third step to rank
alternative plausible rules.

>>>>

A
Suricata rule is divided in three parts---action, header, and rule
options~\cite{suri-rule-format}. The action part appears as the first
word in the rule description. An action denotes the task that needs to
be executed if the rule pattern is satisfied.
In this example, a
message will be sent to system administrators if the rule pattern is
satisfied. The header comes after the action in the rule
description. It restricts the information flow covered by the
rule. For this rule, the header is \CodeIn{tcp \$HOME\_NET any ->
  \$EXTERNAL\_NET any}. It instructs Suricata to inspect \CodeIn{tcp}
traffic flowing from any port in the home network to any other address
outside the home network. The variables \CodeIn{\$HOME\_NET} and
\CodeIn{\$EXTERNAL\_NET} are configurable. The rule options come after
the header. It is a semi-colon-separated sequence of key-value
pairs. Rule options serve to document the analyzed traffic (\eg{},
``msg'', ``classtype'', and ``sid'') and to characterize the attack
pattern (\eg, ``flags'', ``threshold'', and ``content''). \tname{}
produces options of the second kind, which are \emph{relevant} to
capture the attack. For example, the option \CodeIn{content:<arg>} checks if
the argument is present in the payload of the message.

\vspace{-1ex}
\begin{table}[h!]
  \footnotesize
  \caption{\label{table:rules}Example options of Suricata.}
  \vspace{-2ex}
  \centering
  \begin{tabular}{clp{5.5cm}}
    \toprule
    \multicolumn{1}{c}{\#} & \multicolumn{1}{c}{Name} &  \multicolumn{1}{c}{Description}\\
    \midrule     
    1 & dsize & matches on the size of the packet payload\\
    2 & itype &  matches on a specific ICMP type\\
    3 & icode & matches on a specific ICMP code\\
    4 & icmp\_seq  & checks a ICMP sequence number\\
    5 & icmp\_id & matches on specific ICMP id-values\\
    6 & window & checks the size of the TCP window\\
    7 & flags & matches on TCP flags\\
    8 & fragbits & checks if the fragmentation or reserved bits are set in the IP header\\
    9 & threshold & minimum threshold for a rule before it generates alerts\\
    10 & content & checks if argument is present on the payload\\
    \bottomrule
  \end{tabular}
\end{table}
%\vspace{-1ex}

@@

The main challenge in synthesizing rules is to discover the set of
options in a rule that blocks the malicious traffic, but allows the
benign traffic to pass. \tname{} proceeds as follows to synthesize
rules for this attack. First, \tname{} analyzes the malicious traffic and extracts options
expressed in that traffic. Recall from
Section~\ref{sec:rules-and-packets} that rule options can be inferred
from packet contents. In this case, however, that information alone is
insufficient as the malicious traffic is characterized as a sequence
of messages. \tname{} finds the following set of options at this stage by
analyzing the flow and contents of the messages:

%% \Mar{...and where did they come from?} \Luc{Sim, floods podem ser compostos por milhares de pacotes, um dos datasets q encontrei possui mais de 500k pacotes de flood. Os datasets foram encontrados na internet.}

@@

Finally, \tname{} ranks the inferred rules according to heuristics
learned from a database of existing rules. For example, one heuristic
function \tname{} uses for ranking leverages the amount of options a
rule of a given protocol typically has---the score of a rule on this
function is proportional to the popularity of the number of options it
declares. The list below shows the rule options for the top-5 rules
\tname{} produces for this attack.\Mar{algum problema em fazer isto?
  exemplos concretos ajudam muito!} \Luc{No caso desse ataque, a regra inicial possui apenas duas opcoes (sem contar o threshold). Entao existem apenas 3 combinacoes possiveis, dessas 3 apenas 2 sao capazes de isolar o ataque do trafego normal. Portanto, o top-2 eh:
  \begin{figure}[h]
    \vspace{-2ex}
    \lstinputlisting[language=C,numbers=none,frame=none,keywords={window,flags}]{top-synflood.rules}
    \vspace{-3ex}  
  \end{figure}

   }\Mar{Lucas, eu nao entendo porque vc. desconsiderou threshold da
  discussao (afinal, a gente destaca thrshold como relevante no inicio
  desta secao). tambem nao entendo porque window---que a gente disse
  acima que *nao* era relevante---aparece nas duas regras
  geradas. enfim, isto esta inconsistente com nosso discurso.}

To sum up, \tname{} uses the messages that manifest an attack in the
first step to create an overspecific seed rule, it uses the benign
traffic in the second step to create a set of plausible rules from the
seed rule, and it uses existing rules from popular sources in the
third step to rank alternative plausible rules.
